MSSM-Penn-Pitt - ACC and DLPFC - DREAM - FEATURECOUNTS - GRCh38 - CQN - Sparse Model

### 2020-03-26 cmc-rnaseq-de/MSSM-Penn-Pitt_ACC_DLPFC_FEATURECOUNTS_GRCh38_CQN_Sparse.Rmd
remove comp. heavy model selection step

### Model Identification
```{r fxn}
# Function to fit a mixed effect model and return BIC
fitMixedEffectModel <- function(varsToFit, groupingVar, lcpm, cnt, covars){

  # Get design matrix
  designMat = getDesignMatrix(covars[colnames(lcpm), varsToFit, drop=F], Intercept = F)
  designMat = designMat$design[,linColumnFinder(designMat$design)$indepCols]

  # Estimate voom weights
  cnt[is.na(cnt)] = 0
  voomWeights = limma::voom(cnt, design = designMat, plot=F)$weights

  modelBIC = foreach::foreach(i = 1:dim(lcpm)[1],
                              .combine = rbind,
                              .packages = c('lme4', 'dplyr'),
                              .export = c('voomWeights')) %dopar% {

                                # Perform weighted mixed linear models for every gene
                                expr = data.frame(Gx = lcpm[i,])
                                data = cbind(expr,
                                             designMat[rownames(expr),],
                                             Individual_ID = groupingVar)
                                formula = paste(setdiff(colnames(data), c('Gx', 'Individual_ID')), collapse = '+') %>%
                                  paste('Gx ~',.,'+ (1|Individual_ID)', collapse = ' ')

                                weights = voomWeights[i,]
                                BIC = NA
                                tryCatch({
                                  mdl = lmer(formula = formula, 
                                             data = data, 
                                             weights = weights)
                                  BIC = BIC(mdl)
                                }, error = function(e){
                                  BIC = NA
                                })
                              return(data.frame(ensembl_gene_id = rownames(lcpm)[i],
                                                BIC = BIC))
                            }
  return(modelBIC)
  }

# Function to select covariates in a step-wise regression fashion
selectCovarsToAdjust <- function(covarsToSelect, adjustCovars, covars, lcpm, cnt, baseModelBIC){

  varsSelected = c()
  pkgs = c('CovariateAnalysis', 'data.table', 'dplyr', 'plyr',
           'edgeR', 'limma', 'lme4', 'foreach', 'doParallel')
  while(length(covarsToSelect) != 0){
    allModelBIC = foreach(i = 1:length(covarsToSelect),
                          .combine = c,
                          .packages = pkgs,
                          .export = c('fitMixedEffectModel')) %dopar% {

                            # Fit a mixed effect linear model and get all BIC
                            modelBIC = fitMixedEffectModel(c(adjustCovars, covarsToSelect[i]),
                                                           covars[colnames(cnt),'Individual_ID'], 
                                                           lcpm, 
                                                           cnt, 
                                                           covars) %>%
                            plyr::rename(c('BIC' = covarsToSelect[i]))

                            return(list(modelBIC))
                          }
    allModelBIC = plyr::join_all(allModelBIC)

    # Find number of genes that are changed based on BIC
    nChange = allModelBIC %>%
      tidyr::gather(ModelName, Value, -ensembl_gene_id) %>%
      dplyr::left_join(baseModelBIC) %>%
      dplyr::mutate(diff = baseModel - Value) %>%
      dplyr::group_by(ModelName) %>%
      dplyr::summarise(better = sum(diff >= 5), 
                       worse = sum(diff <= -5)) %>%
      dplyr::mutate(nchange = better - worse)

    print(nChange)

    # Select variable
    tmp = nChange %>%
      dplyr::filter(nchange > round(dim(cnt)[1]*0.01)) %>%
      dplyr::top_n(1, nchange)

    if (dim(tmp)[1] == 0){
      covarsToSelect = c()
    } else {
      baseModelBIC = allModelBIC %>%
        dplyr::select(ensembl_gene_id, one_of(tmp$ModelName))
      colnames(baseModelBIC)[2] = 'baseModel'
      varsSelected = c(varsSelected, tmp$ModelName)
      covarsToSelect = setdiff(covarsToSelect, tmp$ModelName)
      print(paste('Variable chosen to be added:',tmp$ModelName))
    }
  }

  return(varsSelected)
}
```
#### Phase I (clinical, ancestry and sample specific technical variables)
```{r phase1}
# Fit a base model with the following covariates
adjust.covars = c('Dx.Tissue')
baseModelBIC = fitMixedEffectModel(adjust.covars, 
                                   COVARIATES[colnames(NEW.COUNTS),'Individual_ID'], 
                                   CQN.GENE_EXPRESSION$E, 
                                   NEW.COUNTS, 
                                   COVARIATES) %>%
  dplyr::rename(baseModel = BIC)

# Iteratively identify covariates in phase I
phase1.covars = c('Institution', 'Reported_Gender', 'PMI', 'ageOfDeath', 'RIN', 'RIN2', 'EV.1', 'EV.2', 'EV.3', 'EV.4', 'EV.5')
vars.selected = selectCovarsToAdjust(phase1.covars, adjust.covars, 
                                     COVARIATES, CQN.GENE_EXPRESSION$E, 
                                     NEW.COUNTS, baseModelBIC)

adjust.covars = c(adjust.covars, vars.selected)
```
Clinical and technical variables selected in phase I are `r paste(adjust.covars, collapse = ', ')`

#### Phase II (batch effects)
```{r phase2}
# Fit a base model with the following covariates
baseModelBIC = fitMixedEffectModel(adjust.covars, 
                                   COVARIATES[colnames(NEW.COUNTS),'Individual_ID'], 
                                   CQN.GENE_EXPRESSION$E, 
                                   NEW.COUNTS, 
                                   COVARIATES) %>%
  dplyr::rename(baseModel = BIC)

# Iteratively identify covariates in phase I
phase2.covars = c('LibraryBatch', 'FlowcellBatch')
vars.selected = selectCovarsToAdjust(phase2.covars, adjust.covars, 
                                     COVARIATES, CQN.GENE_EXPRESSION$E, 
                                     NEW.COUNTS, baseModelBIC)

adjust.covars = c(adjust.covars, vars.selected)
```
Clinical, technical and batch variables selected after phase I and II of model identification are `r paste(adjust.covars, collapse = ', ')`

#### Phase III (RNASeq alignment specific covariates)
```{r phase3}
# Fit a base model with the following covariates
baseModelBIC = fitMixedEffectModel(adjust.covars, 
                                   COVARIATES[colnames(NEW.COUNTS),'Individual_ID'], 
                                   CQN.GENE_EXPRESSION$E, 
                                   NEW.COUNTS, 
                                   COVARIATES) %>%
  dplyr::rename(baseModel = BIC)

# Iteratively identify covariates in phase I
phase3.covars = c("MappedReads", "IntragenicRate", "IntronicRate", "IntergenicRate", "GenesDetected", "rRNARate", "TotalReads")
vars.selected = selectCovarsToAdjust(phase3.covars, adjust.covars, 
                                     COVARIATES, CQN.GENE_EXPRESSION$E, 
                                     NEW.COUNTS, baseModelBIC)

adjust.covars = c(adjust.covars, vars.selected)
```
#### Variables included in the model 
Covariates were added as fixed effects iteratively in three phases if model improvement by BIC criteria was observed in the majority of genes. Clinical, ancestry and sample-specific technical variables were tested for model improvement first. Covariates related to batch effects were next tested for model improvement and finally, RNASeq alignment-specific covariates.

>>>>

#### Model Selection
```{r set_vars}
adjust.covars <- c("Dx.Tissue", "RIN", "IntronicRate")
```

[Code](https://github.com/kelshmo/analysis/blob/f7345adde2052afbc258773f69930df300989bf3/cmc-rnaseq-de/MSSM-Penn-Pitt_ACC_DLPFC_FEATURECOUNTS_GRCh38_CQN_Sparse.Rmd#L579-L737) for model selection and [BIC Criteria](https://www.synapse.org/#!Synapse:syn21543160)

#### Variables included in the model 
Covariates were added as fixed effects iteratively in three phases if model improvement by BIC criteria was observed in the majority of genes. Clinical, ancestry and sample-specific technical variables were tested for model improvement first. Covariates related to batch effects were next tested for model improvement and finally, RNASeq alignment-specific covariates.
